# Exercise 11 Solution {.unnumbered}

# • Solution {.unnumbered}

- Using the {tidyverse} `read_csv()` function, load the "AVONETdataset1.csv" dataset as a "tibble" named **d**.

```{r message=FALSE}
library(tidyverse)
library(skimr)
library(cowplot)
library(broom)
library(infer)
library(mosaic)
library(ggpubr)

f <- "https://raw.githubusercontent.com/difiore/ada-2022-datasets/main/AVONETdataset1.csv"
d <- read_csv(f, col_names = TRUE)
```

- Winnow the dataset to variables of interest.

```{r}
d <- d %>% select(Species1, Family1, Order1, Beak.Length_Culmen, Beak.Width, Beak.Depth, Tarsus.Length, Wing.Length, Tail.Length, Mass, Habitat, Migration, Trophic.Level, Trophic.Niche, Min.Latitude, Max.Latitude, Centroid.Latitude, Range.Size, Primary.Lifestyle)
```

- Do a bit of exploratory data analysis.

```{r}
skim(d)
```

The taxomic variables, along with **Habitat**, **Trophic.Level**, **Trophic.Niche**, and **Migration**, are categorical, although **Migration** is scored using integer factors as "1", "2", or "3". The remaining variables are numeric.

## Challenge 1 {.unnumbered}

### Step 1 {.unnumbered}

- Make boxplots of log(**Mass**) in relation to **Trophic.Level** and **Migration** and convert the variable **Migration** into a factor.

```{r}
d <- d %>% mutate(
  logMass = log(Mass),
  logRS = log(Range.Size),
  logBeak = log(Beak.Length_Culmen),
  logTarsus = log(Tarsus.Length),
  Migration = as.factor(Migration))

p1 <- ggplot(data = d %>% drop_na(Trophic.Level), aes(x = Trophic.Level, y = log(Mass))) + geom_boxplot()

p2 <- ggplot(data = d %>% drop_na(Migration) , aes(x = as.factor(Migration), y = log(Mass))) + geom_boxplot()
plot_grid(p1, p2, nrow = 1)
```

### Step 2 {.unnumbered}

- Run linear models using the `lm()` function to look at the relationship between log(**Mass**) and **Trophic.Level** and between log(**Mass**) and **Migration**.

```{r}
m1 <- lm(logMass ~ Trophic.Level, data = d)
m2 <- lm(logMass ~ Migration, data = d)
```

- Examine the output of the resultant linear models. Is log(**Mass**) associated with either **Trophic.Level** or **Migration** category? That is, in the global test of significance, is the F statistic large enough to reject the null hypothesis of an F value of zero?

```{r}
summary(m1)
summary(m2)
```

In both cases, the global test is significant.

- Given the regression coefficients returned for your **Migration** model, which **Migration** categor(ies) are different than the reference level? What level is the reference level?

In m2, **Migration** levels 2 and 3 are both different from **Migration** level 1, which is the reference level.

- Relevel and assess differences among the remaining pair of **Migration** categories.

```{r}
d <- d %>% mutate(Migration = relevel(Migration, ref = "3"))
m2 <- lm(logMass ~ Migration, data = d)
summary(m2)
```

Based on the releveled model results, **Migration** levels 2 and 3 are also different from one another.

### Step 3 {.unnumbered}

- Conduct a post-hoc Tukey Significant Differences test to also evaluate which **Migration** categories differ "significantly" from one another.

```{r}
# we need an anova object to run a Tukey test
m2 <- aov(logMass ~ Migration, data = d)
posthoc <- TukeyHSD(m2, which = "Migration", conf.level = 0.95)
plot(posthoc)
```

### Step 4 {.unnumbered}

- Use a permutation approach to inference to generate a null distribution of F statistic values for the model of log(**Mass**) in relation to **Trophic.Level** and calculate a p value for your original F statistic.

```{r}
original.F <- aov(logMass ~ Trophic.Level, data = d) %>%
  tidy() %>%
  filter(term == "Trophic.Level") %>%
  pull(statistic)
# show aov results for F statistic and p value for omnibus F test
original.F

# using {infer}
permuted.F <- d %>%
  specify(logMass ~ Trophic.Level) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 1000, type = "permute") %>%
  calculate(stat = "F")

visualize(permuted.F) + shade_p_value(obs_stat = original.F, direction = "greater")

p.value <- permuted.F %>%
  get_p_value(obs_stat = original.F, direction = "greater")

# or...
# using {mosaic}
reps <- 1000
permuted.F <- do(reps) * {
  d %>%
    mutate(Trophic.Level = sample(Trophic.Level)) %>%
    aov(logMass ~ Trophic.Level, data = .) %>%
    tidy() %>%
    filter(term == "Trophic.Level") %>%
    pull(statistic)
}

permuted.F <- as_tibble(permuted.F) %>%
  rename(stat = "result")

p.value <- permuted.F %>%
  get_p_value(obs_stat = original.F, direction = "greater")
p.value
```

## Challenge 2 {.unnumbered}

### Step 1 {.unnumbered}

- Create the following two new variables and add them to AVONET dataset:
  - **Relative beak length**, which you should calculate as the *residual* of log(**Beak.Length_Culmen**) on log(**Mass**).
  - **Relative tarsus length**, which you should calculate as the *residual* of log(**Tarsus.Length**) on log(**Mass**).

```{r}
relBeak <- lm(logBeak ~ logMass, data = d)
relTarsus <- lm(logTarsus ~ logMass, data = d)
d <- d %>% mutate(
  relBeak = relBeak$residuals,
  relTarsus = relTarsus$residuals)
```

### Step 2 {.unnumbered}

- Make a boxplot or violin plot of your new relative tarsus length variable in relation to **Primary.Lifestyle** and of your new relative beak length variable in relation to **Trophic.Niche**

```{r}
p1 <- ggplot(data = d %>% drop_na(Primary.Lifestyle), aes(x=Primary.Lifestyle, y=relTarsus)) +
  geom_boxplot() +
  theme(
    axis.text.x = element_text(angle = 45, hjust=1))

p2 <- ggplot(data = d %>% drop_na(Trophic.Niche), aes(x=Trophic.Niche, y=relBeak)) +
  geom_boxplot() +
  theme(
    axis.text.x = element_text(angle = 45, hjust=1))
plot_grid(p1, p2, nrow = 1)
```

### Step 3 {.unnumbered}

- Run ANOVA analyses to look at the association between geographic range size and the variable **Migration**. You should first drop those observations for which **Migration** is not scored and also look at the distribution of the variable **Range.Size** to decide whether and how it might need to be transformed.

```{r}
migration <- d %>% drop_na(Migration)
histogram(migration$Range.Size)
histogram(migration$logRS)
```

It looks like **Range.Size** should be log transformed.

```{r message=FALSE}
# look at distribution of the data across levels
ggplot(data = migration, aes(x = Migration, y = logRS)) + geom_violin()
```

```{r}
m3 <- lm(logRS ~ Migration, data = migration)
summary(m3)
``` 

- Based on the global model, is range size associated with form of migration? How much of the variance in your measure of range size is associated with **Migration** behavior level?

Yes, the F statistic is significant, although the R2 value suggests that only ~9% of the variation in log(range size) is associated with **Migration** behavior style.

- Given the regression coefficients returned in output of the model, which **Migration** categor(ies) are different than the reference level? What level is the reference level?

In m3, **Migration** levels 1 and 2 are both different from **Migration** level 3, which is the reference level (because that is what we releveled to above).

- Relevel and assess differences among the remaining pair of **Migration** categories.

```{r}
migration <- migration %>% mutate(Migration = relevel(Migration, ref = "1"))
m3 <- lm(logRS ~ Migration, data = migration)
summary(m3)
```

After releveling, we can see that **Migration** levels 1 and 2 are also different from one another.

- Also conduct a post-hoc Tukey Honest Significant Differences test to also evaluate which **Migration** categories differ "significantly" from one another.

```{r}
posthoc <- TukeyHSD(m3, which = "Migration", conf.level = 0.95)
plot(posthoc)
```

### Step 4 {.unnumbered}

- Winnow your original data to just consider birds from the Infraorder "Passeriformes" (song birds).

```{r}
pass <- d %>% filter(Order1 == "Passeriformes")
```

- Run separate one-factor ANOVA analyses to look at the association between [1] relative beak length and **Primary.Lifestyle** and between [2] relative beak length and **Trophic.Level**. In doing so...

  - Make boxplots of response variable by each predictor and by the combination of predictors.

```{r fig.height=6}
p1 <- ggplot(data=pass, aes(x=Primary.Lifestyle, y=relBeak)) +
  geom_boxplot() +
  theme(
    axis.text.x = element_text(angle = 45, hjust=1))
p2 <- ggplot(data=pass, aes(x=Trophic.Level, y=relBeak)) +
  geom_boxplot() +
  theme(
    axis.text.x = element_text(angle = 45, hjust=1))
p3 <- ggplot(data = pass, aes(x=Primary.Lifestyle, y=relBeak)) +
  geom_boxplot() +
  facet_wrap(~Trophic.Level) +
  theme(
    axis.text.x = element_text(angle = 45, hjust=1))
p4 <- ggplot(data = pass, aes(x=Trophic.Level, y=relBeak)) +
  geom_boxplot() +
  facet_wrap(~Primary.Lifestyle) +
  theme(
    axis.text.x = element_text(angle = 45, hjust=1))
plot_grid(plot_grid(p1, p2, nrow = 1), p3, p4, nrow = 3)
```

  - Run linear models for each predictor separately and interpret the model output.

```{r}
m5 <- lm(relBeak ~ Primary.Lifestyle, data = pass)
m6 <- lm(relBeak ~ Trophic.Level, data = pass)
summary(m5)
summary(m6)
```

Relative beak length is significantly associated with both variables, but the R2 value is quite modest.

### Step 5 {.unnumbered}

- Run a two-factor model to look at the association between relative beak length and both **Primary.Lifestyle** and **Trophic.Level** among the passeriforms. Based on the model output, what would you conclude about how relative beak length is related to these two variables?

```{r}
m7 <- lm(relBeak ~ Primary.Lifestyle + Trophic.Level, data = pass)
tidy(m7)
```

### Step 6 {.unnumbered}

- Finally, run an additional two-way model with the same dataset and predictors, but adding the possibility of an interaction term.

```{r}
m8 <- lm(relBeak ~ Primary.Lifestyle + Trophic.Level + Primary.Lifestyle:Trophic.Level, data = pass)
tidy(m8)
```

### Step 7 {.unnumbered}

- Use the `interaction.plot()` function to visualize the interaction between **Primary.Lifestyle** and **Trophic.Level**.

```{r fig.height=6}
par(mfrow=c(2,1))
interaction.plot(
  x.factor = pass$Primary.Lifestyle,
  xlab = "Primary Lifestyle",
  trace.factor = pass$Trophic.Level,
  trace.label = "Trophic Level",
  response = pass$relBeak,
  fun = base::mean, # make sure we use {base} version
  ylab = "Mean Relative Beak Length"
)

interaction.plot(
  x.factor = pass$Trophic.Level,
  xlab = "Trophic Level",
  trace.factor = pass$Primary.Lifestyle,
  trace.label = "Primary Lifestyle",
  response = pass$relBeak,
  fun = base::mean, # make sure we use {base} version
  ylab = "Mean Relative Beak Length"
)
```

## Additional Steps? {.unnumbered}

Let's look at the model of relative beak size ~ trophic level.

Ratio of variances...

```{r}
sd_ratio <- pass %>% group_by(Trophic.Level) %>%
  summarize(sd = sd(relBeak, na.rm = TRUE)) %>% pull(sd)
(sd_ratio <- max(sd_ratio) / min(sd_ratio))
```

Visual checks for normality of observations and residuals within groups... all seem to deviate from normality!

```{r message=FALSE}
p1 <- ggplot(data = pass, aes(x=relBeak)) + geom_histogram() + facet_wrap(~Trophic.Level)
m6data <- tibble(residuals = m6$residuals, Trophic.Level = m6$model$Trophic.Level)
p2 <- ggqqplot(data = m6data, x = "residuals") + facet_wrap(~Trophic.Level)
plot_grid(p1, p2, nrow=2)
```

We can do the same for our model of log(range size) ~ migration level.

Ratio of variances...

```{r}
sd_ratio <- migration %>%
  group_by(Migration) %>%
  summarize(sd = sd(logRS, na.rm = TRUE)) %>% pull(sd)
(sd_ratio <- max(sd_ratio) / min(sd_ratio))
```

Visual checks... again all seem to deviate from normality!

```{r message=FALSE}
p1 <- ggplot(data = migration, aes(x=logRS)) + geom_histogram() + facet_wrap(~Migration)
m3data <- tibble(residuals = m3$residuals, Migration = m3$model$Migration)
p2 <- ggqqplot(data = m3data, x = "residuals") + facet_wrap(~Migration)
plot_grid(p1, p2, nrow=2)
```

```{r include=FALSE}
detach(package:ggpubr)
detach(package:mosaic)
detach(package:infer)
detach(package:broom)
detach(package:cowplot)
detach(package:skimr)
detach(package:tidyverse)
```
